Geometric models reveal behavioral and neural signatures of how naturalistic experiences are transformed into episodic memories

### 2020-08-28 paper/main.tex
added paxton's distinctiveness figure from slack, minor formatting tweaks
detailed pass through paper up until end of discussion section, with major edits to discussion.  still need to go through methods section carefully

\begin{abstract}
Our ongoing subjective experience reflects external sensory information from each moment, along with additional information from our past that we carry with us into that moment.  The blend of memories, knowledge, emotions, goals, and other internal perceptual and mental states that color our subjective experience provides a \textit{context} for interpreting new information and conceptually linking what is happening now with our prior experiences.  Because this contextual information is often person-specific, the subjective experience that each person encodes into their memory is often idiosyncratic, even for shared experiences and sensory perspectives.  We sought to study which aspects of a shared naturalistic experience were preserved or distorted, and how those distortions compared across individuals.  To this end, we developed a geometric framework for mathematically characterizing the subjective conceptual content of dynamic naturalistic experiences.  We model experiences as \textit{trajectories} through word embedding spaces whose coordinates reflect the universe of thoughts under consideration.  We also demonstrate how \textit{memories} may also be modeled as trajectories through the same spaces.  According to this view, encoding an experience into memory entails geometrically distorting or transforming the \textit{shape} of the original experience's trajectory.  This translates qualitative, neuropsychological questions about how we remember naturalistic experiences into quantitive, geometric questions about the spatial configurations of trajectory shapes.  We applied our framework to data collected as participants watched and verbally recounted a television episode while undergoing functional neuroimaging.  We found that the trajectories of participants' recountings of the episode nearly all captured the coarse spatial properties of the original episode's trajectory (i.e., the essential plot points), but participants differed in their memory for fine details.  We also identified a network of brain structures that were sensitive to the shape of the episode's trajectory through word embedding space, and an overlapping network that predicted, at the time of encoding, how people would distort (transform) the episode's trajectory when they recounted the episode later.  Our work provides insights into how our brains distort and transform our ongoing experiences when we encode them into episodic memories.
\end{abstract}

\section*{Introduction}
What does it mean to \textit{remember} something? In traditional episodic memory experiments \citep[e.g., list-learning or trial-based experiments;][]{Murd62a, Kaha96}, remembering is often cast as a discrete and binary operation: each studied item may be separated from the rest of one's experience and singularly labeled as having been recalled or forgotten. More nuanced studies might incorporate self-reported confidence measures as a proxy for memory strength, or ask participants to discriminate between ``recollecting'' the (contextual) details of an experience or having a general feeling of ``familiarity'' \citep{Yone02}. Using well-controlled, trial-based experimental designs, the field has amassed a wealth of information regarding human episodic memory.  However, there are fundamental properties of the external world and our memories that trial-based experiments are not well suited to capture~\citep[for review, also see][]{KoriGold94, HukEtal18}.  First, our experiences and memories are continuous, rather than discrete---isolating a (naturalistic) event from the context in which it occurs can substantially change its meaning.  Second, whether or not the rememberer has precisely reproduced a specific set of words in describing a given experience is nearly orthogonal to how well they were actually able to remember it.  In classic (e.g., list-learning) memory studies, by contrast, the number or proportion of \textit{exact} recalls is often considered to be a primary metric for assessing the quality of participants' memories.  Third, one might remember the \textit{essence} (or a general summary) of an experience but forget (or neglect to recount) particular details.  Capturing the essence of what happened is often a main goal of recounting an episodic memory to a listener, whereas the inclusion of specific, low-level details is often less pertinent.

@@

Occasionally, this gradual ``drift'' of our ongoing experience is punctuated by sudden changes, or ``shifts''~\citep[e.g., when we walk through a doorway; ][]{RadvZack17}.  Prior research suggests that these sharp transitions (termed \textit{event boundaries}) help to discretize our experiences (and their mental representations) into \textit{events}~\citep{RadvZack17, BrunEtal18, HeusEtal18b, ClewDava17, EzzyDava11, DuBrDava13}.  The interplay between the stable (within-event) and transient (across-event) temporal dynamics of an experience also provides a potential framework for transforming experiences into memories that distills those experiences down to their essence.  For example, prior work has shown that event boundaries can influence how we learn sequences of items~\citep{HeusEtal18b, DuBrDava13}, navigate~\citep{BrunEtal18}, and remember and understand narratives~\citep{ZwaaRadv98, EzzyDava11}.  This work also suggests a means of distinguishing the essence of an experience from its low-level details.  The overall structure of events and event transitions reflects how the high-level experience unfolds (i.e., its essence), while subtler event-level properties reflect low-level details.  Prior research has also implicated a network of brain regions (including the hippocampus and the medial prefrontal cortex) in playing a critical role in transforming experiences into structured and consolidated memories ~\citep{TompDava17}.

Here, we sought to examine how the temporal dynamics of a ``naturalistic'' experience were later reflected in participants' memories.  We also sought to leverage the above conceptual insights into the distinctions between an experience's essence and low-level details to build models that explicitly quantified these distinctions.  We analyzed an open dataset that comprised behavioral and functional Magnetic Resonance Imaging (fMRI) data collected as participants viewed and then verbally recounted an episode of the BBC television series \textit{Sherlock}~\citep{ChenEtal17}.  We developed a computational framework for characterizing the temporal dynamics of the moment-by-moment content of the episode, and of participants' verbal recalls.  Specifically, we use topic modeling~\citep{BleiEtal03} to characterize the thematic conceptual (semantic) content present in each moment of the episode and recalls, and hidden Markov models~\citep{Rabi89, BaldEtal17} to discretize this evolving semantic content into events.  In this way, we cast both naturalistic experiences and memories of those experiences as geometric \textit{trajectories} that describe how they evolve over time. Under this framework, successful remembering entails verbally ``traversing'' the content trajectory of the episode, thereby reproducing the shape (essence) of the original experience.  Our framework captures the episode's essence in the sequence of geometric coordinates for its events, and its low-level details by examining its within-event geometric properties.

Comparing the overall shapes of the topic trajectories for the episode and participants' recalls reveals which aspects of the episode's essence were preserved (or discarded) in the translation into memory.  We also develop two metrics for assessing participants' memories for low-level details: (1) the \textit{precision} with which a participant recounts details about each event, and (2) the \textit{distinctiveness} of each recall event, relative to other recalled events.  We examine how these metrics relate to overall memory performance as judged by third-party human annotators.  We also compare and contrast our general approach to studying memory for naturalistic experiences with standard metrics for assessing performance on more traditional memory tasks, such as list-learning.  Last, we leverage our framework to identify networks of brain structures whose responses (as participants watched the episode) reflected the temporal dynamics of either the episode or how participants would later recount it.

@@

\cite{ChenEtal17} collected hand-annotated information about each of 1,000 (manually identified) scenes spanning the roughly 50 minute video used in their experiment.  This information included: a brief narrative description of what was happening, the location where the scene took place, the names of any characters on the screen, and other similar details (for a full list of annotated features, see \textit{Methods}).  We took from these annotations the union of all unique words (excluding stop words, such as ``and,'' ``or,'' ``but,'' etc.) across all features and scenes as the ``vocabulary'' for the topic model.  We then concatenated the sets of words across all features contained in overlapping sliding windows of (up to) 50 scenes, and treated each window as a single ``document'' for the purpose of fitting the topic model.  Next, we fit a topic model with (up to) $K = 100$ topics to this collection of documents.  We found that 32 unique topics (with non-zero weights) were sufficient to describe the time-varying content of the episode (see \textit{Methods}; Figs.~\ref{fig:schematic}, \topics).  Note that our approach is similar in some respects to Dynamic Topic Models~\citep{BleiLaff06}, in that we sought to characterize how the thematic content of the episode evolved over time.  However, whereas Dynamic Topic Models are designed to characterize how the properties of \textit{collections} of documents change over time, our sliding window approach allows us to examine the topic dynamics within a single document (or video).  Specifically, our approach yielded (via the topic proportions matrix) a single \textit{topic vector} for each sliding window of annotations transformed by the topic model.  We then stretched (interpolated) the resulting windows-by-topics matrix to match the time series of the 1,976 fMRI volumes collected as participants viewed the episode.

\begin{figure}[tp]
\centering
\includegraphics[width=1\textwidth]{figs/schematic}
\caption{\small \textbf{Topic weights in episode and recall content.} We used hand-annotated descriptions of each manually identified scene from the episode to fit a topic model.  Three example episode frames (first row) and their associated descriptions (second row) are displayed.  The third row shows an example participant's later recalls of the same three scenes.  We used the topic model (fit to the episode annotations) to estimate topic vectors for each moment of the episode and each sentence of participants' recalls.  Example topic vectors are displayed in the bottom row (blue: episode annotations; green: example participant's recalls).  Three topic dimensions are shown (the highest-weighted topics for each of the three example scenes, respectively), along with the 10 highest-weighted words for each topic.  Figure~\topics~provides a full list of the top 10 words from each of the discovered topics.}
\label{fig:schematic}
\end{figure}
% add an x-label under the "

The 32 topics we found were heavily character-focused (i.e., the top-weighted word in each topic was nearly always a character) and could be roughly divided into themes centered around Sherlock Holmes (the titular character), John Watson (Sherlock's close confidant and assistant), supporting characters (e.g., Inspector Lestrade, Sergeant Donovan, or Sherlock's brother Mycroft), or the interactions between various groupings of these characters (see Fig.~\topics).  This likely follows from the frequency with which these terms appeared in the episode annotations.  Several of the identified topics were highly similar, which we hypothesized might allow us to distinguish between subtle narrative differences if the distinctions between those overlapping topics were meaningful.  The topic vectors for each timepoint were also \textit{sparse}, in that only a small number (typically one or two) of topics tended to be ``active'' in any given timepoint (see Fig.~\ref{fig:model}A).  Further, the dynamics of the topic activations appeared to exhibit \textit{persistence} (i.e., given that a topic was active in one timepoint, it was likely to be active in the following timepoint) along with \textit{occasional rapid changes} (i.e., occasionally topics would appear to spring into or out of existence).  These two properties of the topic dynamics may be seen in the block diagonal structure of the timepoint-by-timepoint correlation matrix (Fig.~\ref{fig:model}B) and reflect the gradual drift and sudden shifts fundamental to the temporal dynamics of many real-world experiences, as well as television episodes.  Given this observation, we adapted an approach devised by \cite{BaldEtal17}, and used a hidden Markov model (HMM) to identify the \textit{event boundaries} where the topic activations changed rapidly (i.e., the boundaries of the blocks in the temporal correlation matrix; event boundaries identified by the HMM are outlined in yellow in Fig.~\ref{fig:model}B).  Part of our model fitting procedure required selecting an appropriate number of ``events'' into which the topic trajectory should be segmented.  To accomplish this, we used an optimization procedure that maximized the difference between the topic weights for timepoints within an event versus timepoints across multiple events (see \textit{Methods} for additional details).  We then created a stable ``summary'' of the content within each episode event by averaging the topic vectors across the timepoints spanned by each event (Fig.~\ref{fig:model}C).

\begin{figure}[tp]
\centering
\includegraphics[width=\textwidth]{figs/eventseg}
\caption{\small \textbf{Modeling naturalistic stimuli and recalls.} All panels: darker colors indicate greater values; range: [0, 1].  \textbf{A.} Topic vectors ($K = 100$) for each of the 1976 episode timepoints.  \textbf{B.} Timepoint-by-timepoint correlation matrix of the topic vectors displayed in Panel A.  Event boundaries discovered by the HMM are denoted in yellow (30 events detected).  \textbf{C.} Average topic vectors for each of the 30 episode events. \textbf{D.} Topic vectors for each of 265 sliding windows of sentences spoken by an example participant while recalling the episode.  \textbf{E.} Timepoint-by-timepoint correlation matrix of the topic vectors displayed in Panel D. Event boundaries detected by the HMM are denoted in yellow (22 events detected).  For similar plots for all participants, see Figure~\corrmats.  \textbf{F.} Average topic vectors for each of the 22 recall events from the example participant.  \textbf{G.} Correlations between the topic vectors for every pair of episode events (Panel C) and recall events (from the example participant; Panel F).  For similar plots for all participants, see Figure~\matchmats.  \textbf{H.} Average correlations between each pair of episode events and recall events (across all 17 participants).  To create the figure, each recalled event was assigned to the episode event with the most correlated topic vector (yellow boxes in panels G and H).}
\label{fig:model}
\end{figure}
% I'm all for citing software packages, but where what's the distinction between packages we do cite (seaborn, brainiak, sklearn, hypertools, quail, umap, etc.) and don't cite (wordcloud, nilearn, fastdtw, etc.)?
% - Paxton
% I also flagged the seaborn reference for removal here. Def not critical to understanding the figure - Andy

Given that the time-varying content of the episode could be segmented cleanly into discrete events, we wondered whether participants' recalls of the episode also displayed a similar structure.  We applied the same topic model (already trained on the episode annotations) to each participant's recalls.  Analogously to how we parsed the time-varying content of the episode, to obtain similar estimates for each participant's recall transcript, we treated each overlapping  window of (up to 10) sentences from their transcript as a ``document,'' and computed the most probable mix of topics reflected in each timepoint's sentences.  This yielded, for each participant, a number-of-windows by number-of-topics topic proportions matrix that characterized how the topics identified in the original episode were reflected in the participant's recalls.  An important feature of our approach is that it allows us to compare participants' recalls to events from the original episode, despite that different participants used widely varying language to describe the events, and that those descriptions often diverged in content and quality from the episode annotations.  This ability to match up conceptually related text that differs in specific vocabulary, detail, and length is an important benefit of projecting the episode and recalls into a shared ``topic'' space.  An example topic proportions matrix from one participant's recalls is shown in Figure~\ref{fig:model}D.

@@

The above results demonstrate that topic models capture the dynamic conceptual content of the episode and participants' recalls of the episode.  Further, the episode and recalls exhibit event boundaries that can be identified automatically using HMMs to segment the dynamic content.  Next, we asked whether some correspondence might be made between the specific content of the events the participants experienced in the episode, and the events they later recalled.  We labeled each recalled event as matching the episode event with the most similar (i.e., most highly correlated) topic vector (Figs.~\ref{fig:model}G, \matchmats).  This yielded a sequence of ``presented'' events from the original episode, and a (potentially differently ordered) sequence of ``recalled'' events for each participant.  Analogous to classic list-learning studies, we can then examine participants' recall sequences by asking which events they tended to recall first~\citep[probability of first recall; Fig.~\ref{fig:list-learning}A;][]{AtkiShif68, PostPhil65, WelcBurn24}; how participants most often transition between recalls of the events as a function of the temporal distance between them~\citep[lag-conditional response probability; Fig.~\ref{fig:list-learning}B;][]{Kaha96}; and which events they were likely to remember overall~\citep[serial position recall analyses; Fig.~\ref{fig:list-learning}C;][]{Murd62a}. Some of the patterns we observed appeared to be similar to classic effects from the list-learning literature.  For example, participants had a higher probability of initiating recall with the first event in the sequence (Fig.~\ref{fig:list-learning}A) and a higher probability of transitioning to neighboring events with an asymmetric forward bias (Fig.~\ref{fig:list-learning}B). However, unlike what is typically observed in list-learning studies, we did not observe patterns comparable to the primacy or recency serial position effects (Fig.~\ref{fig:list-learning}C).  We hypothesized that participants might be leveraging the meaningful narrative associations and references over long timescales throughout the episode.

@@

\begin{figure}[t]
  \centering
  \includegraphics[width=1\textwidth]{figs/list_learning}
  \caption{\small \textbf{Naturalistic extensions of classic list-learning memory analyses.} \textbf{A.} The probability of first recall as a function of the serial position of the event in the episode. \textbf{B}.  The probability of recalling each event, conditioned on having most recently recalled the event \textit{lag} events away in the episode.  \textbf{C.} The proportion of participants who recalled each event, as a function of the serial position of the events in the episode.  All panels: error ribbons denote bootstrap-estimated standard error of the mean.}
  \label{fig:list-learning}
\end{figure}

\begin{figure}[t]
  \centering
  \includegraphics[width=1\textwidth]{figs/precision_distinctiveness}
  \caption{\small \textbf{Novel content-based metrics of naturalistic memory: precision and distinctiveness.} \textbf{A.} The episode-recall correlation matrix for a representative participant (17).  The yellow boxes highlight the maximum correlation in each column.  The example participant's overall precision score was computed as the average across correlation values in the yellow boxes.  Their distinctiveness score was computed as the average (over recall events) of 1 minus the average correlation between each recall event and all other recall events that do not display a box in the same row.  \textbf{B.} The (Pearson's) correlation between precision and hand-annotated memory performance. \textbf{C.} The correlation between distinctiveness and hand-annotated memory performance. \textbf{D.} The correlation between precision and the number of episode events successfully recalled, as determined by our model. \textbf{E.} The correlation between distinctiveness and the number of episode events successfully recalled, as determined by our model.}
  \label{fig:precision-distinctiveness}
\end{figure}

@@

\textit{Precision} is intended to capture the ``completeness'' of recall, or how fully the presented content was recapitulated in memory.  We define a recall event's precision as the maximum correlation between the topic proportions of that recall event and any episode event (Fig.~\ref{fig:precision-distinctiveness}).  In other words, given that a recalled event best matches a particular episode event, more precisely recalled events overlap more strongly with the conceptual content of the original episode event.  When a given event is assigned a blend of several topics, as is often the case (Fig.~\ref{fig:model}), a high precision score requires recapitulating the relative topic proportions during recall.

A second novel metric we developed is \textit{distinctiveness}, which is intended to capture the ``specificity'' of recall.  In other words, distinctiveness quantifies the extent to which a given recalled event reflects the most similar episode event over and above its reflection of other episode events.  Intuitively, distinctiveness is like a normalized variant of our precision metric.  Whereas precision solely measures how much detail about an episode was captured in someone's recall, distinctiveness penalizes details that also pertain to other episode events.  We define the distinctiveness of an event's recall as its precision expressed in standard deviation units with respect to other episode events.  Specifically, for a given recall event, we compute the correlation between its topic vector and that of each episode event.  This yields a distribution of correlation coefficients (one per episode event).  We subtract the mean and divide by the standard deviation of this distribution to $z$-score the coefficients.  The maximum value in this distribution (which, by definition, belongs to the episode event that best matches the given recall event) is the recall event's distinctiveness score.  In this way, recall events that match one episode event far better than all other episode events will receive a high distinctiveness score.  By contrast, a recall event that matches all episode events roughly equally will receive a comparatively low distinctiveness score.

>>>>

\begin{abstract}
Our ongoing subjective experience reflects external sensory information from each moment, along with additional information from our past that we carry with us into that moment.  The blend of memories, knowledge, emotions, goals, and other internal perceptual and mental states that color our subjective experience provides a \textit{context} for interpreting new information and conceptually linking what is happening now with our prior experiences.  Because this contextual information is often person-specific, the subjective experience that each person encodes into their memory is often idiosyncratic, even for shared experiences and sensory perspectives.  We sought to study which aspects of a shared naturalistic experience were preserved or distorted, and how those distortions compared across individuals.  To this end, we developed a geometric framework for mathematically characterizing the subjective conceptual content of dynamic naturalistic experiences.  We model experiences as \textit{trajectories} through word embedding spaces whose coordinates reflect the universe of thoughts under consideration.  We also demonstrate how \textit{memories} may also be modeled as trajectories through the same spaces.  According to this view, encoding an experience into memory entails geometrically preserving, distorting, or transforming the \textit{shape} of the original experience's trajectory.  This translates qualitative neuropsychological questions about how we remember naturalistic experiences into quantitive geometric questions about the spatial configurations of trajectory shapes.  We applied our framework to data collected as participants watched and verbally recounted a television episode while undergoing functional neuroimaging.  We found that the trajectories of participants' recountings of the episode nearly all captured the coarse spatial properties of the original episode's trajectory (i.e., the essential plot points), but participants differed in their memory for low-level details.  We also identified a network of brain structures that were sensitive to the shape of the episode's trajectory through word embedding space, and an overlapping network that predicted, at the time of encoding, how people would distort (transform) the episode's trajectory when they recounted the episode later.  Our work provides insights into how our brains preserve, distort, and transform our ongoing experiences when we encode them into episodic memories.
\end{abstract}

\section*{Introduction}
What does it mean to \textit{remember} something? In traditional episodic memory experiments \citep[e.g., list-learning or trial-based experiments;][]{Murd62a, Kaha96}, remembering is often cast as a discrete binary operation: each studied item may be separated from the rest of one's experience and labeled as having been either recalled or forgotten. More nuanced studies might incorporate self-reported confidence measures as a proxy for memory strength, or ask participants to discriminate between recollecting the (contextual) details of an experience or having a general feeling of familiarity~\citep{Yone02}. Using well-controlled, trial-based experimental designs, the field has amassed a wealth of information regarding human episodic memory~\citep[for review see][]{Kaha12}.  However, there are fundamental properties of the external world and our memories that trial-based experiments are not well suited to capture~\citep[for review, also see][]{KoriGold94, HukEtal18}.  First, our experiences and memories are continuous, rather than discrete---isolating a naturalistic event from the context in which it occurs can substantially change its meaning.  Second, whether or not the rememberer has precisely reproduced a specific set of words in describing a given experience is nearly orthogonal to how well they were actually able to remember it.  In classic (e.g., list-learning) memory studies, by contrast, the number or proportion of \textit{exact} recalls is often considered to be a primary metric for assessing the quality of participants' memories.  Third, one might remember the essence (or a general summary) of an experience but forget (or neglect to recount) particular low-level details.  Capturing the essence of what happened is often a main goal of recounting an episodic memory to a listener, whereas the inclusion of specific low-level details is often less pertinent.

@@

Occasionally, this gradual drift of our ongoing experience is punctuated by sudden changes, or shifts~\citep[e.g., when we walk through a doorway; ][]{RadvZack17}.  Prior research suggests that these sharp transitions (termed \textit{event boundaries}) help to discretize our experiences (and their mental representations) into \textit{events}~\citep{RadvZack17, BrunEtal18, HeusEtal18b, ClewDava17, EzzyDava11, DuBrDava13}.  The interplay between the stable (within-event) and transient (across-event) temporal dynamics of an experience also provides a potential framework for transforming experiences into memories that distills those experiences down to their essences.  For example, prior work has shown that event boundaries can influence how we learn sequences of items~\citep{HeusEtal18b, DuBrDava13}, navigate~\citep{BrunEtal18}, and remember and understand narratives~\citep{ZwaaRadv98, EzzyDava11}.  This work also suggests a means of distinguishing the essence of an experience from its low-level details.  The overall structure of events and event transitions reflects how the high-level experience unfolds (i.e., its essence), while subtler event-level properties reflect low-level details.  Prior research has also implicated a network of brain regions (including the hippocampus and the medial prefrontal cortex) in playing a critical role in transforming experiences into structured and consolidated memories ~\citep{TompDava17}.

Here, we sought to examine how the temporal dynamics of a naturalistic experience were later reflected in participants' memories.  We also sought to leverage the above conceptual insights into the distinctions between an experience's essence and its low-level details to build models that explicitly quantified these distinctions.  We analyzed an open dataset that comprised behavioral and functional Magnetic Resonance Imaging (fMRI) data collected as participants viewed and then verbally recounted an episode of the BBC television show \textit{Sherlock}~\citep{ChenEtal17}.  We developed a computational framework for characterizing the temporal dynamics of the moment-by-moment content of the episode, and of participants' verbal recalls.  Our framework uses topic modeling~\citep{BleiEtal03} to characterize the thematic conceptual (semantic) content present in each moment of the episode and recalls and hidden Markov models~\citep{Rabi89, BaldEtal17} to discretize this evolving semantic content into events.  In this way, we cast both naturalistic experiences and memories of those experiences as geometric \textit{trajectories} that describe how they evolve over time. Under this framework, successful remembering entails verbally traversing the content trajectory of the episode, thereby reproducing the shape (essence) of the original experience.  Our framework captures the episode's essence in the sequence of geometric coordinates for its events, and its low-level details by examining its within-event geometric properties.

Comparing the overall shapes of the topic trajectories for the episode and participants' recalls reveals which aspects of the episode's essence were preserved (or discarded) in the translation into memory.  We also develop two metrics for assessing participants' memories for low-level details: (1) the \textit{precision} with which a participant recounts details about each event, and (2) the \textit{distinctiveness} of each recalled event, relative to other events.  We examine how these metrics relate to overall memory performance as judged by third-party human annotators.  We also compare and contrast our general approach to studying memory for naturalistic experiences with standard metrics for assessing performance on more traditional memory tasks, such as list-learning.  Last, we leverage our framework to identify networks of brain structures whose responses (as participants watched the episode) reflected the temporal dynamics of the episode and/or how participants would later recount it.

@@

\cite{ChenEtal17} collected hand-annotated information about each of 1,000 (manually identified) scenes spanning the roughly 50 minute video used in their experiment.  This information included: a brief narrative description of what was happening, the location where the scene took place, the names of any characters on the screen, and other similar details (for a full list of annotated features, see \textit{Methods}).  We took from these annotations the union of all unique words (excluding stop words, such as ``and,'' ``or,'' ``but,'' etc.) across all features and scenes as the vocabulary for the topic model.  We then concatenated the sets of words across all features contained in overlapping sliding windows of (up to) 50 scenes, and treated each window as a single document for the purpose of fitting the topic model.  Next, we fit a topic model with (up to) $K = 100$ topics to this collection of documents.  We found that 32 unique topics (with non-zero weights) were sufficient to describe the time-varying content of the episode (see \textit{Methods}; Figs.~\ref{fig:schematic}, \topics).  We note that our approach is similar in some respects to Dynamic Topic Models~\citep{BleiLaff06}, in that we sought to characterize how the thematic content of the episode evolved over time.  However, whereas Dynamic Topic Models are designed to characterize how the properties of \textit{collections} of documents change over time, our sliding window approach allows us to examine the topic dynamics within a single document (or video).  Specifically, our approach yielded (via the topic proportions matrix) a single \textit{topic vector} for each sliding window of annotations transformed by the topic model.  We then stretched (interpolated) the resulting windows-by-topics matrix to match the time series of the 1,976 fMRI volumes collected as participants viewed the episode.

\begin{figure}[tp]
\centering
\includegraphics[width=1\textwidth]{figs/schematic}
\caption{\small \textbf{Topic weights in episode and recall content.} We used hand-annotated descriptions of each manually identified scene from the episode to fit a topic model.  Three example episode frames (first row) and their associated descriptions (second row) are displayed.  The third row shows an example participant's recountings of the same three scenes.  We used the topic model (fit to the episode annotations) to estimate topic vectors for each moment of the episode and each sentence of participants' recalls.  Example topic vectors are displayed in the bottom row (blue: episode annotations; green: example participant's recalls).  Three topic dimensions are shown (the highest-weighted topics for each of the three example scenes, respectively), along with the 10 highest-weighted words for each topic.  Figure~\topics~provides a full list of the top 10 words from each of the discovered topics.}
\label{fig:schematic}
\end{figure}
% add an x-label under the "

The 32 topics we found were heavily character-focused (i.e., the top-weighted word in each topic was nearly always a character) and could be roughly divided into themes centered around Sherlock Holmes (the titular character), John Watson (Sherlock's close confidant and assistant), supporting characters (e.g., Inspector Lestrade, Sergeant Donovan, or Sherlock's brother Mycroft), or the interactions between various groupings of these characters (see Fig.~\topics).  This likely follows from the frequency with which these terms appeared in the episode annotations.  Several of the identified topics were highly similar, which we hypothesized might allow us to distinguish between subtle narrative differences if the distinctions between those overlapping topics were meaningful.  The topic vectors for each timepoint were also \textit{sparse}, in that only a small number (typically one or two) of topics tended to be ``active'' in any given timepoint (see Fig.~\ref{fig:model}A).  Further, the dynamics of the topic activations appeared to exhibit \textit{persistence} (i.e., given that a topic was active in one timepoint, it was likely to be active in the following timepoint) along with \textit{occasional rapid changes} (i.e., occasionally topic weights would change abruptly from one timepoint to the next).  These two properties of the topic dynamics may be seen in the block diagonal structure of the timepoint-by-timepoint correlation matrix (Fig.~\ref{fig:model}B) and reflect the gradual drift and sudden shifts fundamental to the temporal dynamics of many real-world experiences, as well as television episodes.  Given this observation, we adapted an approach devised by \cite{BaldEtal17}, and used a hidden Markov model (HMM) to identify the \textit{event boundaries} where the topic activations changed rapidly (i.e., the boundaries of the blocks in the temporal correlation matrix; event boundaries identified by the HMM are outlined in yellow in Fig.~\ref{fig:model}B).  Part of our model fitting procedure required selecting an appropriate number of events into which the topic trajectory should be segmented.  To accomplish this, we used an optimization procedure that maximized the difference between the topic weights for timepoints within an event versus timepoints across multiple events (see \textit{Methods} for additional details).  We then created a stable summary of the content within each episode event by averaging the topic vectors across the timepoints spanned by each event (Fig.~\ref{fig:model}C).

\begin{figure}[tp]
\centering
\includegraphics[width=\textwidth]{figs/eventseg}
\caption{\small \textbf{Modeling naturalistic stimuli and recalls.} All panels: darker colors indicate greater values; range: [0, 1].  \textbf{A.} Topic vectors ($K = 100$) for each of the 1976 episode timepoints.  \textbf{B.} Timepoint-by-timepoint correlation matrix of the topic vectors displayed in Panel A.  Event boundaries discovered by the HMM are denoted in yellow (30 events detected).  \textbf{C.} Average topic vectors for each of the 30 episode events. \textbf{D.} Topic vectors for each of 265 sliding windows of sentences spoken by an example participant while recalling the episode.  \textbf{E.} Timepoint-by-timepoint correlation matrix of the topic vectors displayed in Panel D. Event boundaries detected by the HMM are denoted in yellow (22 events detected).  For similar plots for all participants, see Figure~\corrmats.  \textbf{F.} Average topic vectors for each of the 22 recall events from the example participant.  \textbf{G.} Correlations between the topic vectors for every pair of episode events (Panel C) and recall events (from the example participant; Panel F).  For similar plots for all participants, see Figure~\matchmats.  \textbf{H.} Average correlations between each pair of episode events and recall events (across all 17 participants).  To create the figure, each recalled event was assigned to the episode event with the most correlated topic vector (yellow boxes in panels G and H).}
\label{fig:model}
\end{figure}

Given that the time-varying content of the episode could be segmented cleanly into discrete events, we wondered whether participants' recalls of the episode also displayed a similar structure.  We applied the same topic model (already trained on the episode annotations) to each participant's recalls.  Analogously to how we parsed the time-varying content of the episode, to obtain similar estimates for each participant's recall transcript, we treated each overlapping  window of (up to 10) sentences from their transcript as a document, and computed the most probable mix of topics reflected in each timepoint's sentences.  This yielded, for each participant, a number-of-windows by number-of-topics topic proportions matrix that characterized how the topics identified in the original episode were reflected in the participant's recalls.  An important feature of our approach is that it allows us to compare participants' recalls to events from the original episode, despite that different participants used widely varying language to describe the events, and that those descriptions often diverged in content and quality from the episode annotations.  This ability to match up conceptually related text that differs in specific vocabulary, detail, and length is an important benefit of projecting the episode and recalls into a shared topic space.  An example topic proportions matrix from one participant's recalls is shown in Figure~\ref{fig:model}D.

@@

The above results demonstrate that topic models capture the dynamic conceptual content of the episode and participants' recalls of the episode.  Further, the episode and recalls exhibit event boundaries that can be identified automatically using HMMs to segment the dynamic content.  Next, we asked whether some correspondence might be made between the specific content of the events the participants experienced in the episode, and the events they later recalled.  We labeled each recalled event as matching the episode event with the most similar (i.e., most highly correlated) topic vector (Figs.~\ref{fig:model}G, \matchmats).  This yielded a sequence of ``presented'' events from the original episode, and a (potentially differently ordered) sequence of ``recalled'' events for each participant.  Analogous to classic list-learning studies, we can then examine participants' recall sequences by asking which events they tended to recall first~\citep[probability of first recall; Fig.~\ref{fig:list-learning}A;][]{AtkiShif68, PostPhil65, WelcBurn24}; how participants most often transitioned between recalls of the events as a function of the temporal distance between them~\citep[lag-conditional response probability; Fig.~\ref{fig:list-learning}B;][]{Kaha96}; and which events they were likely to remember overall~\citep[serial position recall analyses; Fig.~\ref{fig:list-learning}C;][]{Murd62a}. Some of the patterns we observed appeared to be similar to classic effects from the list-learning literature.  For example, participants had a higher probability of initiating recall with the first event in the sequence (Fig.~\ref{fig:list-learning}A) and a higher probability of transitioning to neighboring events with an asymmetric forward bias (Fig.~\ref{fig:list-learning}B). However, unlike what is typically observed in list-learning studies, we did not observe patterns comparable to the primacy or recency serial position effects (Fig.~\ref{fig:list-learning}C).  We hypothesized that participants might be leveraging the meaningful narrative associations and references over long timescales throughout the episode.

@@

\begin{figure}[tp]
  \centering
  \includegraphics[width=1\textwidth]{figs/list_learning}
  \caption{\small \textbf{Naturalistic extensions of classic list-learning memory analyses.} \textbf{A.} The probability of first recall as a function of the serial position of the event in the episode. \textbf{B}.  The probability of recalling each event, conditioned on having most recently recalled the event \textit{lag} events away in the episode.  \textbf{C.} The proportion of participants who recalled each event, as a function of the serial position of the events in the episode.  All panels: error ribbons denote bootstrap-estimated standard error of the mean.}
  \label{fig:list-learning}
\end{figure}

\begin{figure}[tp]
  \centering
  \includegraphics[width=1\textwidth]{figs/precision_distinctiveness}
  \caption{\small \textbf{Novel content-based metrics of naturalistic memory: precision and distinctiveness.} \textbf{A.} The episode-recall correlation matrix for a representative participant (17).  The yellow boxes highlight the maximum correlation in each column.  The example participant's overall precision score was computed as the average across correlation values in the yellow boxes.  Their distinctiveness score was computed as the average (over recall events) of the $z$-scored (within column) event precisions. \textbf{B.} The (Pearson's) correlation between precision and hand-counted number of recalled scenes.  \textbf{C.} The correlation between distinctiveness and hand-counted number of recalled scenes. \textbf{D.} The correlation between precision and the number of episode events successfully recalled, as determined by our model. \textbf{E.} The correlation between distinctiveness and the number of episode events successfully recalled, as determined by our model.}
  \label{fig:precision-distinctiveness}
\end{figure}

@@

\textit{Precision} is intended to capture the ``completeness'' of recall, or how fully the presented content was recapitulated in a participant's recounting.  We define a recall event's precision as the maximum correlation between the topic proportions of that recall event and any episode event (Fig.~\ref{fig:precision-distinctiveness}).  In other words, given that a recalled event best matches a particular episode event, more precisely recalled events overlap more strongly with the conceptual content of the original episode event.  When a given event is assigned a blend of several topics, as is often the case (Fig.~\ref{fig:model}), a high precision score requires recapitulating the relative topic proportions during recall.

\textit{Distinctiveness} is intended to capture the ``specificity'' of recall.  In other words, distinctiveness quantifies the extent to which a given recalled event reflects the most similar episode event over and above its reflection of other episode events.  Intuitively, distinctiveness is like a normalized variant of our precision metric.  Whereas precision solely measures how much detail about an episode was captured in someone's recall, distinctiveness penalizes details that also pertain to other episode events.  We define the distinctiveness of an event's recall as its precision expressed in standard deviation units with respect to other episode events.  Specifically, for a given recall event, we compute the correlation between its topic vector and that of each episode event.  This yields a distribution of correlation coefficients (one per episode event).  We subtract the mean and divide by the standard deviation of this distribution to $z$-score the coefficients.  The maximum value in this distribution (which, by definition, belongs to the episode event that best matches the given recall event) is that recall event's distinctiveness score.  In this way, recall events that match one episode event far better than all other episode events will receive a high distinctiveness score.  By contrast, a recall event that matches all episode events roughly equally will receive a comparatively low distinctiveness score.
